Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-09-25 paper/paper.tex
a

Forward and cross-decomposition (CCA, PLS) models are regularly used to identify the causal contribution of collinear features onto multi-dimensional observations (e.g. \citep{naselaris2011encoding}). These approaches typically lead to multiple coefficients for each features (i.e. one per dimension of $Y$ or one per component respectively). Furthermore, these  coefficients can be difficult to summarize into a single 'causal' estimation. By contrast, B2B quick (Fig. \ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending towards 1 and 0 for causal and non-causal features.

A variety of other statistical methods applied to neuroimaging data have been proposed to clarify what is being represented in brain responses - i.e. what feature specifically cause brain activity. One of the popular linear method is Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}, and consists in analyzing the similarity of brain responses associated with specific categorical conditions (e.g. distinct images), by (1) fitting one-against-all classifiers on each condition and (2) testing whether these classifiers can discriminate all other conditions. The resulting confusion matrix is then analyzed in an unsupervised manner to reveal which conditions lead to similar brain activity patterns. B2B differs from RSA in that (1) it uses regressions instead of classifications, and can thus generalize to new items and new contexts and (2) it is fully supervised.

Furthermore, B2B closely relates to CCA \citep{cca_hotelling}, but departs from it in four main aspects. First, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$. By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions. Second, CCA is based a eigenvector decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability. Third, CCA does not typically optimize two distinct regularization parameters. Finally, CCA does not use different data splits to estimate $G$ and $H$. Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Figs. \ref{fig:percondition} and \ref{fig:auc_plots}). Extensions of CCA have been explored in neuroimaging (see \citep{de2019multiway} for brief review), but typically do not address the issue of causal detection.

>>>>

Forward and cross-decomposition (CCA, PLS) models are regularly used to identify the causal contribution of collinear features onto multi-dimensional observations (e.g. \citep{naselaris2011encoding}). These approaches typically lead to multiple coefficients for each features (i.e. one per dimension of $Y$ or one per component respectively). Furthermore, these  coefficients can be difficult to summarize into a single causal estimate. By contrast, B2B quickly (Fig. \ref{fig:duration}) leads to a single unbiased scalar values $\hat E$ tending towards 1 and 0 for causal and non-causal features respectively.

A variety of other statistical methods applied to neuroimaging data have been proposed to clarify what is being represented in brain responses - i.e. what feature causes specific brain activity. One of the popular linear method is Representational Similarity Analysis (RSA) \citep{kriegeskorte2008representational}, and consists in analyzing the similarity of brain responses associated with specific categorical conditions (e.g. distinct images), by (1) fitting one-against-all classifiers on each condition and (2) testing whether these classifiers can discriminate all other conditions. The resulting confusion matrix is then analyzed in an unsupervised manner to reveal which conditions lead to similar brain activity patterns. B2B differs from RSA in that (1) it uses regressions instead of classifications, and can thus generalize to new items and new contexts and (2) it is fully supervised.

Furthermore, B2B closely relates to CCA \citep{cca_hotelling}, but departs from it in four main aspects. First, B2B is not symmetric between $X$ and $Y$: it aims to identify specific causal features by first optimizing over the decoders $G$ and then over $H$. By contrast, CCA is symmetric between $X$ and $Y$, and aims to find $G$ and $H$ such that they project $X$ and $Y$ on maximally correlated dimensions. Second, CCA is based an eigen decomposition of $XH$ and $YG$ - the corresponding canonical components are thus mixing the $X$ features in way that limit interpretability. Third, CCA does not typically optimize two distinct regularization parameters. Finally, CCA does not use different data splits to estimate $G$ and $H$. Together, these differences may explain why B2B reliably outperform CCA on estimating causal influences (Figs. \ref{fig:percondition} and \ref{fig:auc_plots}). Extensions of CCA have been explored in neuroimaging (see \citep{de2019multiway} for brief review), but typically do not address the issue of causal discovery.
